---
title: "facet distributions"
author: "Johan Ravn Cornelius"
date: "2026-05-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source('Brownian_motion.R',keep.source = TRUE)
source('EM.R',keep.source = TRUE)
library(ggplot2)
library(latex2exp)
library(cowplot)
library(tidyverse)
library(extrafont)
```

Double-well potential model, including choice of parameters:
```{r}
y <- -0.05
eps <- 0.1
sigma <- 1.5

F <- function(x){(x-x^3-y)/eps}
DF <- function(x){(1-3*x^2)/eps}
G <- function(x){sigma}
```

Splitting scheme functions:
```{r}
A <- function(c){DF(c)}
btilde <- function(c){ c - F(c)/DF(c) }
N <- function(x,c){(-x^3+3*c^2*x-2*c^3)/eps}
muh <- function(x,h,c){exp(A(c)*h)*x+(1-exp(A(c)*h))*btilde(c)}
Omegah <- function(h,c){  sigma^2 * (exp(2*A(c)*h)-1)/(2*A(c))  }
fh_fun <- function(x, h,c) {
  ODEfun <- function(t, state, parms) {
    u1 <- state[1]
    du1 <- N(u1,c)
    list(c(du1)) }
  out <- deSolve::ode(y = x, times = c(0, h), func = ODEfun, parms = NULL, method = 'lsoda')
  as.numeric(out[nrow(out), -1])
}
```

Number of simulations:
```{r}
M <- 5000
```

Choice of transition density:
```{r}
x0 <- -1
x0_left <- x0
h <- 0.08
```

Simulating EM steps (independent of chosen splitting):
```{r}
hEM <- h/2000
EMtimes <- seq(0,h,by=hEM)
EMsim <- rep(NA,M)
for(i in 1:M){
  EMsteps <-  EM_sim_fast(F,G,x0,EMtimes,n_paths=1,progress = FALSE)[[1]]
  EMsim[i] <- EMsteps[nrow(EMsteps),1]
}
EMdata <- data.frame(EMsim)
EMdata$scheme <- "True distribution (EM)"
EMdata$x0 <- x0
names(EMdata) <- c("sim", "scheme", "Initial value")
```

Finding a c that minimizes bias function:
```{r}
B <- function(b,x,y,eps,sigma){
  denom <- 1/(24*eps^3)
  top <- (-6*b^3+3*y)*x^4 +
    x^3*(-36*b^4-24*sigma^2*eps+36*b^2-8) +
    x^2*(54*(b^3-y)*(b^2-1/3)) +
    x*(-24*b^6+54*b^2*eps*sigma^2+36*b^3*y-12*sigma^2*eps-12*y^2) +
    6*b^5-9*b^4*y+b^3*(-18*sigma^2*eps-4) +
    6*b^2*y+12*eps*sigma^2*y
  return(top*denom)
}
get_roots <- function(y){
  r <- polyroot(c(-y, 1, 0, -1))  # -x^3 + x - y = 0
  r <- Re(r[abs(Im(r)) < 1e-8])  # keep real roots
  sort(r)
}
inv <- function(x,y,sigma,eps){exp(2*(0.5*x^2-0.25*x^4-y*x)/(sigma^2*eps))}
avg_absbias_A1 <- function(b,y,eps,sigma){
  uns_root <- median(get_roots(y))
  Z <- integrate(inv,-Inf,Inf,y=y,sigma=sigma,eps=eps)$value
  p_inv <- 1/integrate(function(x,y,sigma,eps) {inv(x,y,sigma,eps) / Z} , -Inf,uns_root,y,sigma,eps)$value
  integrand <- function(x){
    abs(B(b,x,y,eps,sigma) ) * inv(x,y,sigma,eps) / Z
  }
  abs(integrate(integrand, -Inf, uns_root)$value*p_inv)
}
avg_absbias_A3 <- function(b,y,eps,sigma){
  uns_root <- median(get_roots(y))
  Z <- integrate(inv,-Inf,Inf,y=y,sigma=sigma,eps=eps)$value
  p_inv <- 1/integrate(function(x,y,sigma,eps) {inv(x,y,sigma,eps) / Z} , uns_root,Inf,y,sigma,eps)$value
  integrand <- function(x){
    abs(B(b,x,y,eps,sigma) )* inv(x,y,sigma,eps) / Z
  }
  abs(integrate(integrand, uns_root, Inf)$value*p_inv)
}

c_A1 <- optimize(f=avg_absbias_A1,y=y,eps=eps,sigma=sigma,interval=c(-2,2))$minimum
c_A3 <- optimize(f=avg_absbias_A3,y=y,eps=eps,sigma=sigma,interval=c(-2,2))$minimum
```

```{r}
StrangSim <- function(x0,h,c,M){
  X0 <- rep(x0,M)
  xi <- rnorm(M,0,sqrt(Omegah(h,c)))
  step1 <- mapply(fh_fun,X0,h/2,c)
  step2 <- mapply(muh,step1,h,c)+xi
  step3 <- mapply(fh_fun,step2,h/2,c)
  return(step3)
}
```

Negative stable fixed point:
```{r}
StrangDataFix <- data.frame(StrangSim(x0,h, min(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataFix$scheme <- "Fixed point"
StrangDataFix$x0 <- x0
names(StrangDataFix) <- c("sim","scheme", "Initial value")
```

Minimal mean bias when x<uns:
```{r}
StrangDataMin <- data.frame(StrangSim(x0,h, c2  ,M))
StrangDataMin$scheme <- "Minimal mean bias"
StrangDataMin$x0 <- x0
names(StrangDataMin) <- c("sim","scheme", "Initial value")
```

... and minimal bias in just this one step:
```{r}
local_bias <- function(b,y,eps,sigma){ abs(B(b,x0,y,eps,sigma)) }
c_local <-  optimize(f=local_bias, y=y,eps=eps,sigma=sigma, interval=c(-2,2) )$minimum
StrangDataLocalMin <- data.frame(StrangSim(x0,h, c_local ,M))
StrangDataLocalMin$scheme <- "Minimal local bias"
StrangDataLocalMin$x0 <- x0
names(StrangDataLocalMin) <- c("sim","scheme", "Initial value")
```


Opposite fixed point:
```{r}
StrangDataOpp <- data.frame(StrangSim(x0,h, max(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataOpp$scheme <- "Opposite fixed point"
StrangDataOpp$x0 <- x0
names(StrangDataOpp) <- c("sim","scheme","Initial value")
```

```{r}
StrangDataEachPoint <- data.frame(StrangSim(x0,h, x0  ,M))
StrangDataEachPoint$scheme <- "Local Strang"
StrangDataEachPoint$x0 <- x0
names(StrangDataEachPoint) <- c("sim","scheme","Initial value")
```

Unstable fixed point:
```{r}
StrangDataUns <- data.frame(StrangSim(x0,h, median(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataUns$scheme <- "Unstable fixed point"
StrangDataUns$x0 <- x0
names(StrangDataUns) <- c("sim","scheme","Initial value")
```


Data when x<uns:
```{r}
data <- rbind(EMdata, StrangDataFix
              #,StrangDataMin
              ,StrangDataOpp
               , StrangDataEachPoint
              #, StrangDataLocalMin
           #  , StrangDataUns
              )
```


Plot:
```{r eval=FALSE, include=FALSE}
colors <- c("Fixed point"="red",
            "Minimal mean bias"="green3", 
            "Opposite fixed point"="blue",
            "EM"="black")
ggplot(data, aes(x=sim, color=scheme)) + geom_density(bw=0.17,lwd=1.5)+ geom_vline(xintercept=median(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey",linetype=2)+
  geom_vline(xintercept=min(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey")+
  geom_vline(xintercept=max(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey")+
  geom_vline(xintercept=-x0)+
  labs(x=TeX('$X_h$'),y="Density",color="Distribution",title=TeX('$X_0=-0.5$'))+
    theme_bw(base_size=16)+
  theme(panel.grid.major = element_blank(), # remove the ugly lines
        panel.grid.minor = element_blank())+
  theme(legend.position="right",legend.key.size= unit(1.2, "cm"),)+
  theme(strip.text.x = element_text(size = 16, color = "black", face = "bold"),
        strip.text.y = element_text(size = 16, color = "black", face = "bold"),
        strip.text = element_text(face="bold"),
        strip.background = element_rect(fill="grey")) # facet strips
```



Adding x0>uns to the data:


```{r}
x0 <- 0.6
EMsim2 <- rep(NA,M)
for(i in 1:M){
  EMsteps2 <-  EM_sim_fast(F,G,x0,EMtimes,n_paths=1,progress = FALSE)[[1]]
  EMsim2[i] <- EMsteps2[nrow(EMsteps2),1]
}
EMdata2 <- data.frame(EMsim2)
EMdata2$scheme <- "True distribution (EM)"
EMdata2$x0 <- x0
names(EMdata2) <- c("sim", "scheme", "Initial value")
StrangDataFix2 <- data.frame(StrangSim(x0,h, max(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataFix2$scheme <- "Fixed point"
StrangDataFix2$x0 <- x0
names(StrangDataFix2) <- c("sim","scheme", "Initial value")
StrangDataMinPos2 <- data.frame(StrangSim(x0,h, c2max  ,M))
StrangDataMinPos2$scheme <- "Minimal mean bias"
StrangDataMinPos2$x0 <- x0
names(StrangDataMinPos2) <- c("sim","scheme","Initial value")
StrangDataOpp2 <- data.frame(StrangSim(x0,h, min(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataOpp2$scheme <- "Opposite fixed point"
StrangDataOpp2$x0 <- x0
names(StrangDataOpp2) <- c("sim","scheme","Initial value")
StrangDataEachPoint2 <- data.frame(StrangSim(x0,h, x0  ,M))
StrangDataEachPoint2$scheme <- "Local Strang"
StrangDataEachPoint2$x0 <- x0
names(StrangDataEachPoint2) <- c("sim","scheme","Initial value")
local_bias_2 <- function(b,y,eps,sigma){ abs(B(b,x0,y,eps,sigma)) }
c_local_2 <-  optimize(f=local_bias_2, y=y,eps=eps,sigma=sigma, interval=c(-2,2) )$minimum
StrangDataLocalMin2 <- data.frame(StrangSim(x0,h, c_local ,M))
StrangDataLocalMin2$scheme <- "Minimal local bias"
StrangDataLocalMin2$x0 <- x0
names(StrangDataLocalMin2) <- c("sim","scheme", "Initial value")
StrangDataUns2 <- data.frame(StrangSim(x0,h, median(Re(polyroot(c(-y,1,0,-1))))  ,M))
StrangDataUns2$scheme <- "Unstable fixed point"
StrangDataUns2$x0 <- x0
names(StrangDataUns2) <- c("sim","scheme","Initial value")
```



```{r}
data2 <- rbind(EMdata2,StrangDataFix2,
               # StrangDataMinPos2,
               StrangDataOpp2
               ,StrangDataEachPoint2
               #, StrangDataLocalMin2
              # ,StrangDataUns2
               )
longdata <- rbind(data, data2)
```


Plot 2:
```{r eval=FALSE, include=FALSE}
ggplot(data2, aes(x=sim, color=scheme)) + geom_density(bw=0.22,lwd=1.5)+
  labs(x=TeX('$X_h$'),y="Density",color="Distribution",title=TeX('$X_0=0.2$'))
```

```{r}
names(longdata)<-c("sim","scheme","X_0")
longdata$scheme <- factor(longdata$scheme,levels=c('True distribution (EM)', 'Fixed point', 'Opposite fixed point', 'Minimal mean bias', 'Minimal local bias', 'Local Strang', 'Unstable fixed point'))

appender <- function(string) 
    TeX(paste("$X_0 = $", string)) 

break_func <- function(x){if (mean(x)<0){
  c(x0_left,  -1,   0,1)
}else {
    c(-1, 0, x0, 1)
  }
}
label_func <- function(x){if (mean(x)<0){
  c(TeX('$x_0$'),  -1,  0,1)
}else {
    c(-1, 0, TeX('$x_0$'), 1)
  }
}

p <- ggplot(longdata, aes(x=sim,color=scheme)) + geom_density(bw=0.15,lwd=1.5,key_glyph='path')+
  geom_vline(xintercept=median(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey",linetype=2)+
  geom_vline(xintercept=min(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey")+
  geom_vline(xintercept=max(Re(polyroot(c(-y,1,0,-1)))),col="darkgrey")+
  geom_vline(data=filter(longdata,X_0==x0_left),aes(xintercept=x0_left))+
  geom_vline(data=filter(longdata,X_0==x0),aes(xintercept=x0))+
  scale_x_continuous(breaks=break_func,labels=label_func)+
  facet_wrap(~X_0,scales='free_x',labeller = as_labeller(appender, 
                            default = label_parsed) )+
  scale_color_manual(values=c("gold", "red", "blue","green4"#, "magenta"
                              # ,"purple","brown" 
                              )
                     )+
#  coord_cartesian(xlim=c(-1.2,1.2),ylim=c(0,2))+
  theme_bw(base_size=20   #, base_family = "Times New Roman"
           )+
  theme(axis.text.x=element_text(size=18) )+
  theme(panel.grid.major = element_blank(), # remove the ugly lines
        panel.grid.minor = element_blank())+
  theme(legend.position="right",legend.key.size= unit(1.2, "cm"),)+
  theme(strip.text.x = element_text(size = 16, color = "black", face = "bold"),
        strip.text.y = element_text(size = 16, color = "black", face = "bold"),
        strip.text = element_text(face="bold"),
        strip.background = element_rect(fill="grey")) + # facet strips
  labs(x=TeX('$X_h$'),y="Density",color="Distribution")


p
#ggsave(plot = p, width = 13, height = 6, dpi = 300, filename = "par fra Ex1, x0 -1,3 and 0,6 , h 0,08 , 5k x 2k version 2.pdf")

```



